\documentclass[a4paper, 11pt]{article}

\usepackage{graphicx}
\usepackage{graphics}
\usepackage{verbatim}
\usepackage{listings}
\usepackage{color}
\usepackage{RR,RRthemes}
\usepackage{hyperref}

%%\usepackage[frenchb]{babel} % optionnel
%%
%% date de publication du rapport
\RRdate{Juin 2011}
%%
%% Cas d'une version deux
%% \RRversion{2}
%% date de publication de la version 2
%% \RRdater{Novembre  2006}
\usepackage{listings}

\usepackage{amssymb}
\usepackage{xspace} 
\usepackage{array} 
\usepackage{multirow}
\newcommand\hyph{\nobreak\hskip0pt-\nobreak\hskip0pt\relax}
\newlength\savedwidth
\newcommand\whline{\noalign{\global\savedwidth
  \arrayrulewidth\global\arrayrulewidth 1.5pt}
  \hline \noalign{\global\arrayrulewidth
  \savedwidth}
}
\newcolumntype{I}{!{\vrule width 1.5pt}}
\renewcommand{\arraystretch}{1.5}

\newcommand{\kaapi}{\textsc{X-Kaapi}\xspace}

%%% For all listing figures
\definecolor{MyDarkBlue}{rgb}{0.254901960784314, 0.411764705882353, 0.882352941176471}

\lstset{commentstyle=\color{blue}}
\lstset{language=C}
\lstset{stringstyle=\ttfamily}
\lstset{ classoffset=1
           }
\lstset{ classoffset=2 
           }
\lstset{ classoffset=3
           }
           
\lstset{classoffset=0, showstringspaces=false}

\RRdate{November 2011}

%%
\RRauthor{
Fabien Le Mentec
  \and
Vincent Faucher
  \and
Thierry Gautier
}

\authorhead{Gautier \& Faucher \& Le Mentec}
\RRtitle{\kaapi FORTRAN programming interface}
\RRetitle{\kaapi FORTRAN programming interface}
\titlehead{\kaapi FORTRAN programming interface}

\RRabstract{This report defines the \kaapi FORTRAN programming interface.}

\RRmotcle{}
\RRkeyword{parallel computing, \kaapi, FORTRAN}
\RRprojets{MOAIS}
\RRdomaine{3} % cas du domaine numero 1
\RRthemeProj{moais} % theme du projet Apics
\RCGrenoble % Grenoble - Rh\^one-Alpes

\begin{document}

% \makeRR % cas d'un rapport de recherche
\makeRT % cas d'un rapport technique.

\tableofcontents
\addtocontents{toc}{\protect\setcounter{tocdepth}{1}}

\newpage
\section{Software installation}\label{sec:userinstall}

\kaapi is both a programming model and a runtime for high performance parallelism targeting multicore and distributed architectures. 
It relies on the work stealing paradigm.
\kaapi was developed in the MOAIS INRIA project by Thierry Gautier, Fabien Le Mentec, Vincent Danjean and Christophe Laferrière in the early stage of the library.

In this report, only the programming model based on the C API is presented.
The runtime library comes also with a full set of complementary programming interfaces: C, C++ and STL-like interfaces. The C++ and STL interfaces, at a higher level than the C interface, may be directly used for developing parallel programs or libraries.

\subsubsection*{Supported Platforms}
\kaapi targets essentially SMP and NUMA platforms. The runtime should run
on any system providing:
\begin{itemize}
\item a GNU toolchain (4.3),
\item the pthread library,
\item Unix based environment.
\end{itemize}
It has been extensively tested on the following operating systems:
\begin{itemize}
\item GNU-Linux with x86\_64 architectures,
\item MacOSX/Intel processor.
\end{itemize}

There is no version for Windows yet.

\subsubsection*{\kaapi Contacts}
If you wish to contact the XKaapi team, please visite the web site at:
\begin{center}
\url{http://kaapi.gforge.inria.fr}
\end{center}

\newpage
\section{Initialization and termination}

\subsection{Synopsis}
\begin{small}
\lstset{commentstyle=\color{blue}}
\lstset{language=C}
\begin{lstlisting}[frame=tb]
SUBROUTINE KAAPIF_INIT(INTEGER*4 FLAGS)
SUBROUTINE KAAPIF_FINALIZE()
\end{lstlisting}
\end{small}

\subsection{Description}
\paragraph{}
\textit{KAAPIF\_INIT} initializes the runtime. It must be called once per program
before using any of the other routines. If successful, there must be a
corresponding \textit{KAAPIF\_FINALIZE} at the end of the program.

\subsection{Parameters}
\begin{itemize}
\item \textit{FLAGS}: if not zero, start only the main thread to avoid disturbing
the execution until tasks are actually scheduled. The other threads are suspended
waiting for a parallel region to be entered (refer to KAAPIF\_BEGIN\_PARALLEL).
\end{itemize}

\subsection{Return value}
\paragraph{}
None.

\subsection{Example}
\paragraph{}
Refer to examples/kaapif/foreach\\
\begin{small}
\begin{lstlisting}[frame=tb]
  PROGRAM MAIN
    CALL KAAPIF_INIT(1)

    ...

    CALL KAAPIF_FINALIZE()
  END PROGRAM MAIN
\end{lstlisting}
\end{small}


\newpage
\section{Concurrency}

\subsection{Synopsis}
\begin{small}
\lstset{commentstyle=\color{blue}}
\lstset{language=C}
\begin{lstlisting}[frame=tb]
INTEGER*4 FUNCTION KAAPIF_GET_CONCURRENCY()
INTEGER*4 FUNCTION KAAPIF_GET_THREAD_NUM()
\end{lstlisting}
\end{small}

\subsection{Description}
\paragraph{}
Concurrency related routines.

\subsection{Return value}
\paragraph{}
\textit{KAAPIF\_GET\_CONCURRENCY} returns the number of parallel thread available
to the \kaapi runtime.
\paragraph{}
\textit{KAAPIF\_GET\_THREAD\_NUM} returns the current thread identifier. Note it
should only be called in the context of a \kaapi thread.

\subsection{Example}
\paragraph{}
Refer to examples/kaapif/foreach


\newpage
\section{Performance}

\subsection{Synopsis}
\begin{small}
\lstset{commentstyle=\color{blue}}
\lstset{language=C}
\begin{lstlisting}[frame=tb]
REAL*8 FUNCTION KAAPIF_GET_TIME()
\end{lstlisting}
\end{small}

\subsection{Description}
\paragraph{}
Capture the current time. Used to measure the time spent in a code region.

\subsection{Parameters}
\paragraph{}
None.

\subsection{Return value}
\paragraph{}
The current time, in microseconds.

\subsection{Example}
\paragraph{}
Refer to examples/kaapif/foreach\\

\begin{small}
\begin{lstlisting}[frame=tb]
  PROGRAM MAIN
    REAL*8 START
    REAL*8 STOP

    CALL KAAPIF_INIT(1)
    START = KAAPIF_GET_TIME()

    ...

    STOP = KAAPIF_GET_TIME()
    CALL KAAPIF_FINALIZE()

    WRITE(*, *) STOP - START

  END PROGRAM MAIN
\end{lstlisting}
\end{small}


\newpage
\section{Independent loops}

\subsection{Synopsis}
\begin{small}
\lstset{commentstyle=\color{blue}}
\lstset{language=C}
\begin{lstlisting}[frame=tb]
SUBROUTINE KAAPIF_FOREACH
(
  BODY,
  INTEGER*4 FIRST, INTEGER*4 LAST,
  INTEGER*4 NARGS,
  ...
)

SUBROUTINE KAAPIF_FOREACH_WITH_FORMAT
(
  BODY,
  INTEGER*4 FIRST, INTEGER*4 LAST,
  INTEGER*4 NARGS,
  ...
)
\end{lstlisting}
\end{small}

\subsection{Description}
\paragraph{}
Those routines run a parallel loop over the range [\textit{FIRST}, \textit{LAST}]
(note this is an \textbf{inclusive} interval).
The loop body is defined by \textit{BODY} whose arguments are given
in parameters. It must have the following prototype:\\
\begin{small}
\lstset{language=C}
\begin{lstlisting}[frame=tb]
SUBROUTINE BODY(I, J, TID, ...)
\end{lstlisting}
\end{small}
\begin{itemize}
\item $[I, J]$ the subrange to process (note that interval is inclusive)
\item \textit{TID} the thread identifier
\end{itemize}

\subsection{Parameters}
\begin{itemize}
\item \textit{BODY}: the function body to be called at each iteration
\item \textit{FIRST}, \textit{LAST}: the iteration range indices, inclusive.
\item \textit{NARGS}: the argument count
\item \textit{...}: the arguments passed to \textit{BODY}. For
\textit{KAAPIF\_FOREACH\_WITH\_FORMAT}, refer to the \textit{KAAPIF\_SPAWN}
documentation.
\end{itemize}

\subsection{Return value}
\paragraph{}
None.

\subsection{Example}
Refer to examples/kaapif/foreach\\
Refer to examples/kaapif/foreach\_with\_format

\begin{small}
\begin{lstlisting}[frame=tb]
  ! computation task entry point
  SUBROUTINE OP(I, J, TID, ARRAY)
    DO K = I, J
      ! process ARRAY(K)
      ...
    END DO
    RETURN
  END

  PROGRAM MAIN
    ...
    ! apply the OP routine on ARRAY[1:SIZE]
    CALL KAAPIF_FOREACH(OP, 1, SIZE, 1, ARRAY)
    ...
  END PROGRAM MAIN
\end{lstlisting}
\end{small}


\newpage
\section{Dataflow programming}
\subsection{Synopsis}
\begin{small}
\lstset{commentstyle=\color{blue}}
\lstset{language=C}
\begin{lstlisting}[frame=tb]
SUBROUTINE KAAPIF_SPAWN
(
  BODY,
  INTEGER*4 NARGS,
  ...
)
\end{lstlisting}
\end{small}

\subsection{Description}
\paragraph{}
Create a new computation task implemented by the function \textit{BODY}.

\paragraph{}
\textit{BODY} is called with the user specified arguments, there is no
argument added by XKAAPI:\\
\begin{small}
\lstset{language=C}
\begin{lstlisting}[frame=tb]
SUBROUTINE BODY(ARG0, ARG1, ...)
\end{lstlisting}
\end{small}

\paragraph{}
Each task parameter is described by 4 successive arguments including:
\begin{itemize}
\item the argument \textit{VALUE},
\item the parameter \textit{TYPE},
\item the element \textit{COUNT},
\item the access \textit{MODE}.
\end{itemize}

\paragraph{}
\textit{TYPE} is one of the following:
\begin{itemize}
\item KAAPIF\_TYPE\_CHAR=0,
\item KAAPIF\_TYPE\_INT=1,
\item KAPAIF\_TYPE\_REAL=2,
\item KAPAIF\_TYPE\_DOUBLE=3.
\end{itemize}

\paragraph{}
If a parameter is an array, \textit{COUNT} must be set to the array size.
For a scalar value, it must be set to 1.

\paragraph{}
\textit{MODE} is one of the following:
\begin{itemize}
\item KAAPIF\_MODE\_R=0 for a read access,
\item KAAPIF\_MODE\_W=1 for a write access,
\item KAAPIF\_MODE\_RW=2 for a read write access,
\item KAAPIF\_MODE\_V=3 for a parameter passed by value.
\end{itemize}

\subsection{Parameters}
\begin{itemize}
\item \textit{BODY}: the task body.
\item \textit{NARGS}: the argument count.
\item \textit{...}: the \textit{VALUE}, \textit{TYPE}, \textit{COUNT}, \textit{MODE} tuple list.
\end{itemize}

\subsection{Return value}
\paragraph{}
None.

\subsection{Example}
\paragraph{}
Refer to examples/kaapif/dfg

\begin{small}
\begin{lstlisting}[frame=tb]
  ! computation task entry point
  SUBROUTINE OP(A, B)
    ! task user specific code
    ...
    RETURN
  END

  PROGRAM MAIN
    ...

    ! spawn a task implemented by the OP routine
    call kaapif_spawn(fu, 2,
    ! argument[0]
     &                42
     &                KAAPIF_TYPE_DOUBLE,
     &                1,
     &                KAAPIF_MODE_V,
    ! argument[1]
     &                42,
     &                KAAPIF_TYPE_DOUBLE,
     &                1,
     &                KAAPIF_MODE_V)

    ...
  END PROGRAM MAIN
\end{lstlisting}
\end{small}


\newpage
\section{Parallel regions}

\subsection{Synopsis}
\begin{small}
\lstset{commentstyle=\color{blue}}
\lstset{language=C}
\begin{lstlisting}[frame=tb]
SUBROUTINE KAAPIF_BEGIN_PARALLEL()
SUBROUTINE KAAPIF_END_PARALLEL(INTEGER*4 FLAGS)
\end{lstlisting}
\end{small}

\subsection{Description}
\paragraph{}
\textit{KAAPIF\_BEGIN\_PARALLEL} and \textit{KAAPIF\_END\_PARALLEL} mark the
start and the end of a parallel region. Regions are used to wakeup and suspend
the \kaapi system threads so they avoid disturbing the application when idle.
This is important if another parallel library is being used. Wether threads
are suspendable or not is controlled according by the KAAPIF\_INIT parameter.

\subsection{Parameters}
\begin{itemize}
\item \textit{FLAGS}: if zero, an implicit synchronization is inserted before
leaving the region.
\end{itemize}

\subsection{Return value}
\paragraph{}
None.

\subsection{Example}
\paragraph{}
Refer to examples/kaapif/dfg\\
\begin{small}
\begin{lstlisting}[frame=tb]
  PROGRAM MAIN
    ...

    CALL KAAPIF_BEGIN_PARALLEL()

    ...

    CALL KAAPIF_END_PARALLEL(1)

    ...
  END PROGRAM MAIN
\end{lstlisting}
\end{small}


\newpage
\section{Synchronization}

\subsection{Synopsis}
\begin{small}
\lstset{commentstyle=\color{blue}}
\lstset{language=C}
\begin{lstlisting}[frame=tb]
SUBROUTINE KAAPIF_SCHED_SYNC()
\end{lstlisting}
\end{small}

\subsection{Description}
\paragraph{}
Synchronize the sequential with the parallel execution flow. When this routine
returns, every computation task has been executed and memory is consistent for
the processor executing the sequential flow.

\subsection{Return value}
\paragraph{}
None.

\subsection{Example}
\paragraph{}
Refer to examples/kaapif/dfg\\
\begin{small}
\begin{lstlisting}[frame=tb]
  PROGRAM MAIN
    ...

    CALL KAAPIF_SYNC()

    ...
  END PROGRAM MAIN
\end{lstlisting}
\end{small}


\end{document}
